Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2019-03-11 Manuscript/manuscript.tex
Update text

Binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,melcr18,nmrlipids_proj4},
which is the case also here as the sodium binding affinity to
POPS lipid bilayer is stronger in Lipid17 simulation with Dang ions than in
ECC-POPS simulation (Fig. \ref{fig:POPS-counterions-dens}).
The different cation binding affinities to lipid bilayers have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:chemstruct_pc_ps} for the labeling) \cite{catte16,melcr18,nmrlipids_proj4}.
According to the ''electrometer concept'' \citep{seelig87}, these order parameters decrease
proportionally to the bound charge and can therefore be used to quantify the cation binding
affinity to lipid bilayers. Because the response of PS lipid headgroup order parameters to the bound charge
is less well undestood and addition of cation may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}, the cation binding affinity to bilayers with
PS lipids has evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,nmrlipids_proj4}.

The situation is more complicated for lipid bilayers containing
PS lipids, for which there is no ion-free reference state due to the presence of counterions.
Moreover, additional ions often lead to phase transitions in such bilayers \cite{nmrlipids_proj4,feigenson86,mattai89,roux91,roux90}

Unlinke in pure PC bilayers,
Interpreting the changes of the head group order parameters of 
mixed PC:PS lipid bilayers 
from both simulations and experiments
is complicated by the presence of counterions. \cite{nmrlipids_proj4}
In order to see the effects of different counterions, 
we performed simulations with \ce{Na+} and also with \ce{K+} counterions, 
which bind more weakly to phospholipids. \cite{nmrlipids_proj4, melcr18}
This can also be seen in the density plots of the counterions in Fig.~\ref{fig:POPS-counterions-dens},
which show a significantly reduced concentration of \ce{K+} ions 
at the membrane head group region compared to \ce{Na+}
for both PS and mixed PC:PS (5:1) bilayers.

%According to the electrometer concept,
%the head group order parameters $\alpha$ and $\beta$ of PC lipids
%change proportionally to the amount of bound charge
%in a fixed ratio. \cite{seelig87,scherer87,roux90} 
%This was demonstrated for a bound positive charge in simulations with POPC and 
%cationic surfactants or aquaeous cations employing the ECC-POPC model \cite{melcr18}.
%In this work, we will use the electrometer concept also for a bound negative charge 
%using PS lipids as negatively charged surfactants. \cite{roux90}

@@

Generally improved behaviour 
of the POPC and POPS head group order parameters 
with \ce{NaCl} or \ce{KCl} concentrations 
was achieved through the combination of models 
ECC-lipids and ECC-ions \citep{martinek17, kohagen16, Pluharova2014}. 
Simulations with these models 
reveal a good agreement with the NMR experiments 
for both neutral and negatively charged membranes.
The results are plotted for \ce{NaCl} in Fig. \ref{fig:delta_ordPar_NaCl_PCPS}, 
and for \ce{KCl} in Fig. \ref{fig:delta_ordPar_KCl_PCPS}.

The slopes of the response of the PC head group order parameters $\alpha$ and $\beta$ 
in the mixed PC:PS (5:1) bilayer with \ce{K+} counterions
(Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix})
agree qualitatively well with experiments for both employed models,
i.e., for both Lipid14/17 and ECC-lipids. 
The same property from the simualtions with \ce{Na+} counterions, 
which were used in the experimental measurement,
are not as good, however.
In particular,
ECC-lipids have a weaker response of their PC head group order parameters compared to the experiments, 
while the simulations with Lipid17 model reveal a completely opposite trend.
This effect likely arises from an overestimated binding of the sodium cations to the phospholipids \cite{nmrlipids_proj4},
which is likely still present in a small amount also in the simulations with ECC-lipids. 
The large difference between the affinities of the employed lipid models to the counterions 
is visible in the density plots in Fig~\ref{fig:POPS-counterions-dens}.

The order parameters of the PS head groups including their changes with the content of PC lipids
are challenging to capture correctly in classical MD simulation models. \cite{nmrlipids_proj4}
From such a perspective,
the changes of the PS head group order parameters from ECC-lipids 
with \ce{Na+}, and also \ce{K+}, counterions 
agree comparably well to the experimental measurements
(Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}). 
Interestingly enough, 
although the Lipid17 model performed well with \ce{K+} counterions 
for the PC head group order parameter changes,
it does not agree in the equivalent changes in PS head groups
pointing at non-negligible effects of even relatively weakly binding cations 
on the structure of negatively charged lipids.

Despite the great improvements in the overall structure of ECC-POPS 
and its response to PC lipids and counterions,
there is still a large room for improving it
to agree with the experimental measurements.

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    The head group order parameters of a POPC:POPS (5:1) bilayer as a function of PS content
    with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions from simulations with ECC-lipids and Lipid17 \cite{lipid17-future} 
    and experiments (only \ce{Na^+} counterions, but shown also in bottom plots). \cite{roux90}. 
    The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
\todo{JOE+SAMULI: split this figure into separate plots for PC and PS, 
and change the PS-plot x-axis to reflect the dilution of POPS by PC (now shows the content of PS). 
The experimental lines in the bottom plot shall be different to make clear that they are not with K counterions (gray dashed?).}
\end{figure*}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer 
    as a function of \ce{NaCl} concentration in the buffer 
    from simulations with different force fields and experiments at 298 K. \citep{roux90}
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \todo{SAMULI: Use the same y-axis scale in the bottom figures showing angles for POPC and POPS to ease the comparison.}
  \end{figure}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_na_k_cl_wat_phos_models-compar_5-6_NaCl-and-KCl-series.pdf}
  \caption{\label{fig:nacl-dens_PCPS} 
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} and \ce{Cl^-} along the membrane normal axis 
    for the negatively charged membranes with the composition of 5\,PC:1\,PS using ECC-lipids and ECC-ions
    from simulations with added concentrations of \ce{KCl} or \ce{NaCl}.  
    The top    profile shows the simulation with an additional \ce{KCl} concentration and \ce{Na^+} counterions. 
    The bottom profile shows the simulation with an additional \ce{NaCl} concentration and \ce{Na^+} counterions, which are not distinguished from the added salt. 
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
  }
\end{figure}

>>>>

Binding affinity of monovalent ions to lipid bilayers depends strongly on
force field parameters \cite{catte16,melcr18,nmrlipids_proj4},
which is the case also here as the sodium binding affinity to
POPS lipid bilayer is stronger in Lipid17 simulation with Dang ions than in
ECC-POPS simulation (Fig. \ref{fig:POPS-counterions-dens}).
The cation binding affinities to lipid bilayers in simulations have been previously evaluated
against experiments using the order parameters of $\alpha$ and $\beta$ carbons in the
phosphatidylcholine (PC) lipid headgroup (see Fig.~\ref{fig:chemstruct_pc_ps} for the labeling) \cite{catte16,melcr18,nmrlipids_proj4},
which decrease proportionally to the bound charge according to the ''electrometer concept'' \citep{seelig87}.
Althought the response of PS lipid headgroup order parameters to the bound charge is also systematic,
it is less well undestood and the addition of cations may cause phase transitions in negatively charged
bilayers \cite{feigenson86,mattai89,roux91,roux90}. Therefore, the cation binding affinity to bilayers with
PS lipids has been evaluated measuring the PC headgroup order parameters from POPC:POPS (5:1)
mixtures\cite{roux90,nmrlipids_proj4}.
Following our previous work \cite{nmrlipids_proj4}, we evaluate the counterion binding affinity
to lipid bilayer in simulations containing PS using two different approaches.

% The situation is more complicated for lipid bilayers containing
%PS lipids, for which there is no ion-free reference state due to the presence of counterions.

%Unlinke in pure PC bilayers,
%Interpreting the changes of the head group order parameters of 
%mixed PC:PS lipid bilayers 
%from both simulations and experiments
%is complicated by the presence of counterions. \cite{nmrlipids_proj4}

@@

%This was demonstrated for a bound positive charge in simulations with POPC and 
%cationic surfactants or aquaeous cations employing the ECC-POPC model \cite{melcr18}.
%In this work, we will use the electrometer concept also for a bound negative charge 
%using PS lipids as negatively charged surfactants. \cite{roux90}

First, we monitor the response of PC headgroup order parameters to the increasing amount of PS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:POPS-counterions-dens}), which
compensates the negative charge of PS lipids at the interface.
\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPS_kcl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix} 
    The head group order parameters of a POPC:POPS (5:1) bilayer as a function of PS content
    with \ce{Na^+} (top) and \ce{K^+} (bottom) counterions from simulations with ECC-lipids and Lipid17 \cite{lipid17-future} 
    and experiments (only \ce{Na^+} counterions, but shown also in bottom plots). \cite{roux90}. 
    The underestimated response of ECC-lipids with \ce{Na^+} counterions 
    is likely due to a still slightly overestimated binding affinity of \ce{Na^+} to the phospholipids,
    which is corroborated by the series with \ce{K^+} counterions (lower affinity than \ce{Na^+}),
    where the response is closer to the experiment, which uses only \ce{Na^+} counterions. 
  }
  \todo{JOE+SAMULI: split this figure into separate plots for PC and PS, 
    and change the PS-plot x-axis to reflect the dilution of POPS by PC (now shows the content of PS). 
    The experimental lines in the bottom plot shall be different to make clear that they are not with K counterions (gray dashed?).} \\
  \todo{Error bar of 0.02 should be shown for the experimental POPS results because they come from different experiments, see caption of figure 8 in NMRlipids IV publication.}
\end{figure*}

Second, we monitor the response of PC headgroup order parameters to the additional
amount of sodium and potassium in POPC:POPS (5:1) mixtures.
In experiments, the addition of K$^+$ ions led to a very modest decrease of the order parameters,
while the decrease was more pronounced with Li$^+$ ions and strongest with divalent Ca$^{2+}$ and Mg$^{2+}$ ions \cite{roux90}.
According to the electrometer concept, the decrease of PC lipid head group order parameters
correlates with the amount of bound cations \cite{akutsu81,altenbach84,seelig87,roux90,catte16,nmrlipids_proj4},
thus the binding affinity of the ions to POPC:POPS (5:1) mixtures increase in order K$^{+}$ $<$ Li$^{+}$  $<$ Mg$^{2+}$  $<$ Ca$^{2+}$.
The response of PC headgroup order parameters to the increasing amount of KCl is close to experiments
in both force fields (Fig. \ref{fig:delta_ordPar_KCl_PCPS}), athought the change of the $\alpha$-carbon
order parameter may be slightly overestimated by the Lipid17/Dang simulations. This could be
explained by the slightly deeper penetration of potassium into the bilayer (Fig. \ref{fig:POPS-counterions-dens}).
Because experimental data with the additional amount of sodium in POPC:POPS (5:1) mixture was not available,
we compared the simulation results to the experimental data with the additional amounts of
lithium and potassium ions, which should give the lower and upper bounds to the sodium binding
affinity (Fig. \ref{fig:delta_ordPar_KCl_PCPS}). The PC headgroup response ot the additional
amount of sodium is clearly overestimated by the Lipid17/Dang simulation, which is in line
the higher binding affinity (Fig. \ref{fig:POPS-counterions-dens}). Also, the response
in ECC-POPS simulations is probably too close to the experimental results for lithium,
indicating slight overbinding also in this simulation. In conclusion, the response of PC
headgroups to added amount of PS and to the added amount of ions suggest that the Lipid17/Dang
simulations sinficantly overestimate the sodium binding affinity to the PS containing lipid bilayers, while
ECC-POPS simulations give results closer to the experiments, but may still slightly overestimate
the binding affinity. The binding affinity of potassium is better described by both force fields.

%Generally improved behaviour 
%of the POPC and POPS head group order parameters 
%with \ce{NaCl} or \ce{KCl} concentrations 
%was achieved through the combination of models 
%ECC-lipids and ECC-ions \citep{martinek17, kohagen16, Pluharova2014}. 
%Simulations with these models 
%reveal a good agreement with the NMR experiments 
%for both neutral and negatively charged membranes.
%The results are plotted for \ce{NaCl} in Fig. \ref{fig:delta_ordPar_NaCl_PCPS}, 
%and for \ce{KCl} in Fig. \ref{fig:delta_ordPar_KCl_PCPS}.

%The order parameters of the PS head groups including their changes with the content of PC lipids
%are challenging to capture correctly in classical MD simulation models. \cite{nmrlipids_proj4}
%From such a perspective,
%the changes of the PS head group order parameters from ECC-lipids 
%with \ce{Na+}, and also \ce{K+}, counterions 
%agree comparably well to the experimental measurements
%(Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}). 
%Interestingly enough, 
%although the Lipid17 model performed well with \ce{K+} counterions 
%for the PC head group order parameter changes,
%it does not agree in the equivalent changes in PS head groups
%pointing at non-negligible effects of even relatively weakly binding cations 
%on the structure of negatively charged lipids.

%Despite the great improvements in the overall structure of ECC-POPS 
%and its response to PC lipids and counterions,
%there is still a large room for improving it
%to agree with the experimental measurements.

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPC_nacl.pdf} 
  \includegraphics[height=\figheight]{../img/ecc_pops/order_parameters_changes_ecc-lip_L14_A-B-PN-COO_POPS_nacl.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PCPS} 
    Changes of the head group order parameters $\alpha$, $\beta$ and the orientations of the carboxylate group and the P-N vector  
    of POPC (left) and POPS (right) phospholipids in a POPC:POPS 5:1 bilayer 
    as a function of \ce{NaCl} concentration in the buffer 
    from simulations with different force fields and experiments at 298 K. \citep{roux90}
    Because data with \ce{NaCl} are not available for POPC, 
    we show experimental data for \ce{LiCl} (dashed line, left) 
    as an upper bound for the magnitude of the response to \ce{NaCl}, 
    which has a lower affinity to phospholipid bilayers compared to \ce{LiCl} \citep{roux90}. 
    The orientation of the \ce{COO^-} group is defined as 
    the connector from the $\beta$ carbon to the carbon in \ce{COO^-} (stars, bottom right). 
  }
  \todo{We should plot the experimental results with KCl also in this figure.} \\
  \todo{Why there are no error bars for ECC-POPS in this and other figures?} \\
  \todo{SAMULI: Use the same y-axis scale in the bottom figures showing angles for POPC and POPS to ease the comparison.}
  \end{figure}

@@

\begin{figure}[tbp!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/density_profiles_na_k_cl_wat_phos_models-compar_5-6_NaCl-and-KCl-series.pdf}
  \caption{\label{fig:nacl-dens_PCPS} 
    Number density profiles of \ce{K^{+}}, \ce{Na^{+}} and \ce{Cl^-} along the membrane normal axis 
    for the negatively charged membranes with the composition of 5\,PC:1\,PS using ECC-lipids and ECC-ions
    from simulations with added concentrations of \ce{KCl} or \ce{NaCl}.  
    The top    profile shows the simulation with an additional \ce{KCl} concentration and \ce{Na^+} counterions. 
    The bottom profile shows the simulation with an additional \ce{NaCl} concentration and \ce{Na^+} counterions, which are not distinguished from the added salt. 
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
  }
  \todo{I am not sure if we need this figure at all.}
\end{figure}
